Gaia identifies VFTS682 as the most massive runaway star known to date

### 2018-05-11 VFTS682.tex
text polishing

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, lead to speculations whether such extreme stars can form in relative isolation. 
%
We use the unprecedented proper motions from the second Gaia data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving with a relative
velocity of about $64\pm21\kms$ away from the star cluster R136.  The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This would make VFTS 682 the most massive runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

\section{Introduction}
\label{sec:intro}

@@

% feels like unnecessary here...
% In this section we summarize the remarkable properties derived for VFTS682 \todo{?}\Secref{data:vfts683}
% in earlier studies and we discuss we Gaia data in \Secref{data:gaia}.

@@

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred a peculiar extinction
($R_V\simeq4.7$), leading to a surprisingly high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citep{schneider:18} estimated
the present-day mass of VFTS682 at  $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonnsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11, kohler:15}. These values place VFTS682 around the
``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

@@

Further worth noticing is the variability of the star.
\citep{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band of the $\sim$10\% level on a timescale of years.
The comment that this is unusual for Wolf-Rayet and more reminiscent
of Luminous Blue Variable (LBV) stars.

\todo{Report mass loss rate.  Report LOS velocity.}
\citep{parker:93}

@@

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS682. }
%
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-45 $\pm$ 25 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  \tablecomments{The peculiar radial
    velocity $\delta v_\mathrm{rad}$ is obtained as the difference
    between the average radial velocity of the 30 Doradus region
    ($270\pm10\kms$) minus the radial velocity measured from the HeII $\lambda4686$
    line for VFTS682 ($315\pm15\kms$).}
    \end{center}
  \label{tab:vfts682}
\end{table}

@@

The Gaia Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}.%  \todo{why do we care about the
%   median? I'm
%   selecting data with errors smaller than this, and I'm not using
%   fainter stars!I think this is not useful here} The median uncertainty in the proper motion
% components the corresponding uncertainties are $0.05\,\masyr$ for
% bright sources  ($G<14$ mag) and $0.2\masyr$ at $G=17\,\mathrm{mag}$
% \citep{lindegren:18}.  The uncertainties quoted above correspond to $xx\kms$ and $yy\kms$ respectively.  This unprecedented accuracy allows the identification of bright runaway stars at the distance to the LMC, 50\,kpc.

VFTS682 is identified with the id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are reliable. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

@@

We define a local standard frame of reference to derive the peculiar velocity
of VFTS682 by selecting from the Gaia DR2 catalog a sample of nearby
stars, following closely the approach of \cite{lennon:18}.
We select all the stars in a target of 0.2 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
not reduces further our sample.

@@

These can be converted in the
the components of the relative transverse velocity $\delta v_\mathrm{RA}=38\pm17\,\kms$,
$\delta v_\mathrm{DEC}=26\pm19\,\kms$, assuming a distance of
50\,kpc. % (we do not account for the uncertainty in the distance
% estimate when propagating errors).
The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the three-dimensional
peculiar speed of the star\footnote{We implicitly assume that the use
  of a different reference frame for $\delta v_\mathrm{rad}$ does not
  introduce significant errors.}:

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec} = \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 64 \pm 21 
  \, \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.6\pm\,0.2 \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =46\pm
17\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}, and we use the approximation $1 \kms \simeq 1
\mathrm{pc \ Myr^{-1}}$. As in the rest of this study, we neglect for
simplicity the error on the distance estimates.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/main_plot_good}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region, starting from the present day position of
    the star. The orange arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards to illustrate the uncertainty on the origin of the
    star. The blue stars
    are a subset of those we use to define the local rest
    frame. \todo{load color picture on background, clip away outliers}}

@@

\todo{check \cite{banerjee} and rewrite next paragraph, they
  have stuff} Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what is their mass
distribution. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.

>>>>

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, lead to speculations whether such extreme stars can form in relative isolation. 
%
We use the unprecedented proper motions from the second Gaia data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving away from the star
cluster R136 with a relative
projected two-dimensional speed of about $\sim$$40\pm20 \kms$velocity,
implying a three-dimensional speed of about $\sim$$60\pm20\kms$ when
accounting for the velocity component along the line of sight.  The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This would make VFTS 682 the most massive runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

\todo{recheck all numbers}
\section{Introduction}
\label{sec:intro}

@@

@@

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred an extinction
of $R_V\simeq4.7$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citep{schneider:18} estimated
the present-day mass of VFTS682 at  $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonnsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11, kohler:15}. These values place VFTS682 around the
``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

@@

Further worth noticing is the variability of the
star. \citet{parker:93} reported B and V magnitudes roughly 0.5\,dex
lower than what found by \citet{evans:11}. \citet{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band of the $\sim$10\% level on a timescale of years.
They commented that this is unusual for Wolf-Rayet and more reminiscent
of Luminous Blue Variable (LBV) stars.

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind hiding the stellar surface.

@@

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS682. }
%
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-30 $\pm$ 20 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  % \tablecomments{}
    \end{center}
  \label{tab:vfts682}
\end{table}

@@

The Gaia Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}. VFTS682 is identified with the id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are reliable. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

@@

We define a local standard frame of reference to derive the peculiar velocity
of VFTS682 by selecting from the Gaia DR2 catalog a sample of nearby
stars, following closely the approach of \cite{lennon:18}.
We select all the stars in a target of 0.2 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
not reduces further our sample. Finally, we remove clear outliers from
the proper motion distribution

@@

These can be converted in the
the components of the relative transverse velocity $\delta v_\mathrm{RA}=38\pm17\,\kms$,
$\delta v_\mathrm{DEC}=26\pm19\,\kms$, assuming a distance of
50\,kpc. % (we do not account for the uncertainty in the distance
% estimate when propagating errors).
The radial velocity from
\cite{bestenlehner:11} then gives the third component along
the line of sight, allowing us to calculate the three-dimensional
peculiar speed of the star:

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec} = \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2+\left(\delta
      v_\mathrm{rad}\right)^2} = 55 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.6\pm\,0.2\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =46\pm
18\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}.% , and we use the approximation $1 \kms \simeq 1
% \mathrm{pc \ Myr^{-1}}$.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/main_plot_good}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region, starting from the present day position of
    the star. The orange arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards (dashed) to illustrate the uncertainty on the origin of the
    star. The blue stars
    are a subset of those we use to define the local rest
    frame. \todo{load color picture on background}}

@@

\todo{check \cite{banerjee:12} and rewrite next paragraph, they
  have stuff} Unfortunately, their simulations did not include stars more
massive than $100\,M_\odot$, so it is difficult to predict, based on
the existence of VFTS682, how many other stars should have been
ejected from the cluster already and what is their mass
distribution. A comprehensive study of the kinematic properties of the
large sample of stars with extremely large inferred masses visible
around R136 is encouraged.
